---
title: "Adaptive_preferential_sampling_with_simulated_data"
author: "Lorenzo Cappello"
date: "August 18, 2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adaptive_preferential_sampling_with_simulated_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Install also the R package adapref at https://github.com/lorenzocapp/adapref. 
This package complements phylodyn allowing for adaptive preferential sampling
```{r setup}
library(adapref)
```

We will first define a sampling rate and the effective population size functions
Note: you need to have installed phylodyn, spmrf and rstan. 

```{r}

set.seed(123)
n<-100

bottleneck_traj<-function(t){
  result=rep(0,length(t))
  result[t<=0.15]<-10
  result[t>0.15 & t<0.3]<-10*exp(3.004268-20*t[t>0.15 & t<0.3]) #50 times smaller
  result[t>=0.3]<-0.5
  return(result)
}

sampling_traj<-function(t){
  result=rep(0,length(t))
  result[t<=0.22]<-10*exp(1.404268-20*t[t<=0.22])
  result[t>0.22]<-0.5
  return(result*40)}
```

We now sample the sampling times, coalescent times, and a corresponding genealogy
Note that in the samp_times vector, we artificially make the first sampling time zero.
n_sampled is going to be a vector of 1, because the Poisson process samples one event a time

We the nplot the simulated genealogy. 
```{r}
samp_times<-c(0,sampsim_thin(max.n=n-1,traj=sampling_traj,xlim=c(0,3))[-n+1])
n_sampled <- rep(1,n) 
simulation<-phylodyn::coalsim(samp_times,n=n_sampled, traj = bottleneck_traj)
tree1<-phylodyn::sample_genealogy(simulation)
treeP<-ape::read.tree(text=tree1$Newick)
plot(treeP)

```

We now define the various representation of the data that are needed for the estimation.
We also define the grid of ngr points. 
```{r}
phyy<-phylodyn::summarize_phylo(treeP)
coaldat<-list(coal_times=phyy$coal_times,intercoal_times=c(phyy$coal_times[1],diff(phyy$coal_times)),samp_times=phyy$samp_times,n_sampled=phyy$n_sampled)
ngr=100
sgrid <- spmrf::makeGrid(coal_times = coaldat$coal_times, samp_times = phyy$samp_times, Ngrid=(ngr+1))
cdat<-samp_and_coalescent_data(samp_times=phyy$samp_times,n_sampled = phyy$n_sampled,coal_times=coaldat$coal_times,grid=sgrid$grid)

```

We can either use INLA or HMC posterior approximation.
First we show how to do it with INLA: the first line use the built in procedure to choose precision hyperparameters of the fields; second line, it is the actual posterior approximation


```{r}

zeta<-set_prec_parameters(phylo = coaldat, ncell = ngr, alpha = 0.05, order.c = 1, order.s = 1)
res<-BNPR_PS_ada(coaldat,lengthout=100,zeta1=zeta$zeta1,zeta2=zeta$zeta2)

```

Now, we plot the posterior of N_e(t) and beta(t)
```{r}
 phylodyn::plot_BNPR(res,heatmaps=TRUE,heatmap_labels = FALSE,ylab="",xlab = "",main="effective pop size")

plot_beta(res,heatmaps=TRUE,heatmap_labels=FALSE,main="beta",ylim=c(10,100),xlim=c(1,0))

```
