---
title: "Count simulated coalescent trees"
author: "Lorenzo Cappello"
date: "7/05/2019"
output: html_document
---

This vignette shows how to simulate data under the infinite sites assumption and then use the counting methodology described in Cappello and Palacios (2019)

We start by loading `ape`, `rPython` and `plyr` (to be installed if not done beforehand)

```{r message=FALSE}
set.seed(1+8)
library("rPython")
library("ape")
library("plyr")
library("phylodyn")

```

Load python function file
```{r}
path <- paste(system.file(package="phylodyn"), "python/Fsamp_4.py", sep="/")
python.load(path)
```

Decide sample size (n), mutation rate (mu)

```{r}
n=10
mu=10
```

Sample a coalescent tree via `rcoal` in the package `ape`

```{r}
tree1<-rcoal(n,tip.label = NULL, br = "coalescent")
```

Superimpose on tree1 mutations as defined by the infinite sites assumptions (Watterson,1975)
```{r}
data1<-simulate_data(mu,tree1)
```
Note: sequences are unlabelled. Labels are given in the later steps

Compute the Tajima perfect phylogeny. 
```{r}
oldsuff<-sufficient_stats(data1)
```

The output `oldsuff` is a Rlist. 
$nodes is a matrix describing the perfect phylogeny tree
1st column: child node label
2nd column: parent node label
3rd column: #elements in the child node
4th column: #identical copies of that node (e.g. there are multiple singletons)
5th column: leaf node (1) or internal node (0)

$mylist include for each row of the matrix a sublist
  $x it is identical to the 4th column
  $y #mutation (if x=1) or vector with #mutations (if x>1)
  



Choose number of importance sampling samples
```{r}
N=50
```



Monte Carlo iteration 
```{r}
est<-siscount_tree(n,mu,N,oldsuff)
```


Compute the estimate for the four different resolutions considered. In the list one can find 
`countTaj`: MonteCarlo estimate #ranked tree shapes
`countKing`: MonteCarlo estimate #ranked labeled trees
`countLT`: MonteCarlo estimate #labeled tree shapes
`countTS`: MonteCarlo estimate #tree shapes

```{r}
print(est$countTaj)
print(est$countKing)
print(est$countLT)
print(est$countTS)
```

It includes also the diagnostics discussed in the paer. For example for the Tajima algorihm

```{r}
print(est$cv2Taj)
print(est$essTaj)
print(est$qTaj)
```


##References

Cappello, L., & Palacios, J. A. (2019). Sequential importance sampling for multi-resolution Kingman-Tajima coalescent counting. arXiv preprint arXiv:1902.05527.










