---
title: "Count simulated coalescent trees"
author: "Lorenzo Cappello"
date: "7/05/2019"
output: html_document
---

This vignette shows how to simulate data under the infinite sites assumption and then use the counting methodology described in Cappello and Palacios (2019)

We start by loading `ape`, `rPython` and `plyr` (to be installed if not done beforehand)

```{r message=FALSE}
set.seed(1+8)
library("rPython")
library("ape")
library("plyr")

```

We also need to load the files with the R functions required for simulation and counting (to be downloaded from phyodyn\R and phylodyn\src)

```{r}
source("function countingv6.R")
source("Tajima.R")
python.load("Fsamp4.py")
```

Decide sample size (n), mutation rate (mu)

```{r}
n=10
mu=10
```

Sample a coalescent tree via `rcoal` in the package `ape`

```{r}
tree1<-rcoal(n,tip.label = NULL, br = "coalescent")
```

Superimpose on tree1 mutations as defined by the infinite sites assumptions (Watterson,1975)
```{r}
data1<-simulate_data(mu,tree1)
```
Note: sequences are unlabelled. Labels are given in the later steps

Compute the Tajima perfect phylogeny. 
```{r}
oldsuff<-sufficient_stats(data1)
```

The output `oldsuff` is a Rlist. 
$nodes is a matrix describing the perfect phylogeny tree
1st column: child node label
2nd column: parent node label
3rd column: #elements in the child node
4th column: #identical copies of that node (e.g. there are multiple singletons)
5th column: leaf node (1) or internal node (0)

$mylist include for each row of the matrix a sublist
  $x it is identical to the 4th column
  $y #mutation (if x=1) or vector with #mutations (if x>1)
  

Compute the Kingman perfect phylogeny.   

```{r}
nodesTaj<-oldsuff$nodes 
nodes<-kingman_perphylo(nodesTaj)
```

Prepare list of active sampling nodes for the Kingman perfect phylogeny.
```{r}
sampnodes<-king_initial_sampnodes(nodes) #sampling nodes
```
and we attach random labels to each sequence
```{r}
labels<-label_king(nodes,sampnodes)
```

Choose number of importance sampling samples
```{r}
N=5000
```

Initialization for the importance sampling cycle
```{r}
t<-rep(1,n-1)
saveTaj<-c()
saveKing<-c()
saveLT<-c()
saveTS<-c()
```

Monte Carlo iteration 
```{r}
for (i in 1:N){
    #1st we sample Tajima tree
    result <- python.call("F_sample", oldsuff) 
    #Compute sampling probability over all possible nodes (in sublist [[1]])
    #Sublist [[2]] and [[3]] give extra information not necessary for this step.
    where_save2<-python.call("calcPF", result$change_F, result$F_nodes, result$family_size, oldsuff,t,"True")
    #compute the statistics here 1/q(g)
    saveTaj[i]<-1/where_save2[[1]]
    
    #As a by product we compute the tree shape statistics
    TS<-tree_shape_discount(result)
    saveTS[i]<-saveTaj[i]/TS
    
    #Now we sample a Kingman topology
    kinglist<-kingtree_sample_v3(nodes,n,sampnodes,labels)
    #compute the statistics here 1/q(g)
    saveKing[i]<-1/prod(kinglist$prob)
    
    #As a by product we compute the  labelled tree shape statistics
    LT<-labelled_trees_discount(kinglist)
    saveLT[i]<-c(saveKing[i]/LT)
    
  }
```


Compute the estimate for the four different resolutions considered: 
`countTaj`: MonteCarlo estimate #ranked tree shapes
`countKing`: MonteCarlo estimate #ranked labeled trees
`countLT`: MonteCarlo estimate #labeled tree shapes
`countTS`: MonteCarlo estimate #tree shapes

```{r}
countTaj<-sum(saveTaj)/N
countKing<-sum(saveKing)/N
countLT<-sum(saveLT)/N
countTS<-sum(saveTS)/N
print(countTaj)
print(countKing)
print(countLT)
print(TS)
```

Finally, we can compute diagnostics for each of the resolution. 

```{r}
diagTaj<-diagnotics(saveTaj,N)
diagKing<-diagnotics(saveKing,N)
diagLT<-diagnotics(saveLT,N)
diagTS<-diagnotics(saveTS,N)
```


##References

Cappello, L., & Palacios, J. A. (2019). Sequential importance sampling for multi-resolution Kingman-Tajima coalescent counting. arXiv preprint arXiv:1902.05527.










