---
title: "Example of SLT inference from simulation"
author: "Julia Palacios, Mackenzie Simper"
output: rmarkdown::html_vignette
date: "2024-10-07"
vignette: >
  %\VignetteIndexEntry{Example_SLT_Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Estimation of mutation rate
===================================

We will first simulate a genealogy with 50 tips under the coalescent model with *exp_tra* effective population size.


```{r}
library("ape")
library("phylodyn")
library("expm")
n<-50
exp_traj<-function(t){
  result=rep(0,length(t))
  result[t<=0.1]<-5
  result[t>0.1 & t<0.5]<-5*exp(0.5-5*t[t>0.1 & t<0.5]) #50 times smaller
  result[t>=0.5]<-0.6766764
  return(result)
}


simul1<-coalsim(samp_times = 0, n_sampled = n, traj = exp_traj,method="tt",val_upper=11)
mytree<-generate_newick((simul1))$newick
library("hash")
set.seed(123)
S = 3 #number of cut sites
theta = matrix(1, nrow = S, ncol = S)
#set different rates for multi-cut
theta[1, 2] <-theta[1, 3] <-theta[2,3]<- 0.5
sdata<-simulate_SLData(numSim=1,mytree,S=3,M=20,theta=theta)[[1]][[2]][1:n,]

```

We now estimate the mutation rates by method of moments

```{r}
t<-coalescent.intervals(mytree)$total.depth
mu_hat<-estimate_mu(sdata,t)
print(mu_hat)


```

Estimation of coalescent times 
=====================================

We will first simulate 5 datasets with the same mutation rates as before but with a smaller genealogy of $n=15$ cells:

```{r}
n<-5
simul1<-coalsim(samp_times = 0, n_sampled = n, traj = exp_traj,method="tt",val_upper=11)
mytree<-generate_newick((simul1))$newick

t<-coalescent.intervals(mytree)$total.depth
numSim=5
bigdata<-simulate_SLData(numSim=numSim,mytree,S=3,M=20,theta=theta)
##just observed alleletypes

datalist<-list()
for (j in 1:numSim){
  datalist[[j]]<-bigdata[[j]][[2]][1:n,]
}

```

We will now assume known mutation rates and tree topology and will estimate coalescent times and $Ne(t)$. In this example we assume all datasets are generated with the same mutation rates

```{r}
thetas<-list()
for (j in 1:numSim){
  thetas[[j]]<-theta
}
library("phangorn")
library("parallel")
times_estimates<-inference_times(Dlist=datalist, thetas=thetas, M = 20,tree=mytree)
plot(cumsum(coalescent.intervals(mytree)$interval.length),cumsum(times_estimates$mle_times))
  
```



Estimation of topology and times
==============================================================

```{r}

all_estimates<-inference_all(Dlist=datalist, t0=t, thetas=thetas, M = 20)
plot(cumsum(coalescent.intervals(mytree)$interval.length),cumsum(times_estimates$mle_times))
  
```

Estimation of Ne
==============================================

```{r}
bound<-t
Ne_est<-MLE_BoundCoal(data=all_estimates$tree,lengthout=4,bound=t,eps=.02,eta=.01)
Ne_oracle<-BNPR(mytree,lengthout=4)
plot_BNPR(Ne_oracle)
points(Ne_est$x,Ne_est$effpop,type="S",col="red")
```
