---
title: "Effective population size inference under Beta-coalescent"
author: "Julie Zhang"
date: "2024-07-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
list.of.packages <- c("ape", "TreeTools", "matrixStats","tidyverse","parallel")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

lapply(list.of.packages, library, character.only = TRUE) 

library(INLA)
library(phylodyn)
set.seed(1234)
```

## Generating a genealogy from the Beta-coalescent

We can generate a multifurcating genealogy with $n=50$ tips from the Beta-coalescent with parameter $\alpha=1.5$ and constant effective population size.This function parallels \texttt{rcoal()} in \texttt{ape}. 

```{r}
n <- 50
alpha <- 1.5
sim_tree <- rbeta_coal(n, alpha)
print(sim_tree$Nnode)
plot(sim_tree, show.tip.label=FALSE)
```

The following example generates a heterochronous genealogy with 50 tips distributed according to the Beta-coalescent with $\alpha=1.5$ and variable effective population size trajectory (exponential trajectory) and specified sampling times (25 sequences at time 0 and 25 sequences at time 1). This function parallels \texttt{coalsim()} implemented in \texttt{phylodyn}. 

```{r}
sim_exp_tree_list <- beta_coalsim(samp_times=c(0,1), n_sampled = c(25,25), alpha=1.5, traj=exp_traj)
sim_exp_tree <- generate_newick_multif(sim_exp_tree_list)$newick
print(sim_exp_tree$Nnode)
plot(ladderize(sim_exp_tree), show.tip.label=FALSE)
```

## Reconstructing effective population size

We now estimate the variable effectve population size trajectory from a given multifurcating genealogy non-parametrically assuming we know the value of $\alpha$. For our example, we will use the genealogy just generated.

```{r}
BNPR_sim_exp <-  BNPR_Lambda(sim_exp_tree,alpha_multif=1.5)
plot_BNPR(BNPR_sim_exp, traj=exp_traj)
```

## Inference Methods

We can first compute the block-size MLE which does not use any information about the branch lengths and apply BNPR. 

```{r pressure, echo=FALSE}
sim_alpha_bs <- beta_coal_block_size_alpha_mle(sim_exp_tree)
BNPR_block_size <- BNPR_Lambda(sim_exp_tree, alpha_multif=sim_alpha_bs)
```

Then, we can apply the empirical Bayes approach until convergence of $\alpha$. 

```{r}
sim_hybrid <- posterior_beta_coal_hybrid(sim_exp_tree)
sim_alpha_eb <- sim_hybrid$alpha_opt
BNPR_hybrid <- sim_hybrid$BNPR 
```

Finally, we apply MCMC with sHMC instead of BNPR for speed considerations. We still recommend use of BNPR in single cases of inference. We use 10\% burn-in out of 5,000 total samples for this vignette, but the default is 20,000

```{r}
sim_MCMC <- suppressWarnings(posterior_beta_coal_MCMC_w_sHMC(sim_exp_tree, nsamp_total=5000))
  
```

Then we can plot the results on one graph 

```{r}
plot_BNPR(BNPR_sim_exp, main=expression(paste('True ', alpha, ': 1.5')))
lines(BNPR_sim_exp$x, BNPR_sim_exp$effpop975, col='gray20', lwd=2, lty=2)
lines(BNPR_sim_exp$x, BNPR_sim_exp$effpop025, col='gray20', lwd=2,lty=2)
lines(BNPR_sim_exp$x, BNPR_block_size$effpop, col='coral1', lwd=2)
lines(BNPR_sim_exp$x, BNPR_block_size$effpop975, col='coral1', lwd=2, lty=2)
lines(BNPR_sim_exp$x, BNPR_block_size$effpop025, col='coral1', lwd=2,lty=2)
lines(BNPR_sim_exp$x, BNPR_hybrid$effpop, col='darkgreen', lwd=2)
lines(BNPR_sim_exp$x, BNPR_hybrid$effpop025, col='darkgreen', lwd=2, lty=2)
lines(BNPR_sim_exp$x, BNPR_hybrid$effpop975, col='darkgreen', lwd=2, lty=2)
lines(sim_MCMC$x, sim_MCMC$effpop, col='royalblue', lwd=2)
lines(sim_MCMC$x, sim_MCMC$effpop025, col='royalblue', lwd=2, lty=2)
lines(sim_MCMC$x, sim_MCMC$effpop975, col='royalblue', lwd=2, lty=2)
traj_vals  <- exp_traj(seq(0,  max(BNPR_sim_exp$grid), by=0.1))
lines(seq(0, max(BNPR_sim_exp$grid), by=0.1),traj_vals , col='purple', lwd=2)
legend('topleft', legend=c("True alpha", 'BS MLE', 'Hybrid', 'MCMC', 'True trajectory'),
       col=c("black", "coral1", 'darkgreen', 'royalblue', 'purple'), lty=1, cex=0.8, lwd=2)

```

We can also calculate how well each method did in reconstructing $N_e(t)$ by computing the coverage, bias, deviance, and MSE of the inferred trajectories versus the true trajectory. 

```{r}
bias <- c(bias_new(BNPR_sim_exp, exp_traj)$avg, bias_new(BNPR_block_size, exp_traj)$avg, 
            bias_new(BNPR_hybrid, exp_traj)$avg, bias_new(sim_MCMC, exp_traj)$avg)
dev <- c(dev_new(BNPR_sim_exp, exp_traj)$avg, dev_new(BNPR_block_size, exp_traj)$avg, 
          dev_new(BNPR_hybrid, exp_traj)$avg, dev_new(sim_MCMC, exp_traj)$avg)
coverage <- c(envelope_new(BNPR_sim_exp, exp_traj)$avg, envelope_new(BNPR_block_size, exp_traj)$avg, 
              envelope_new(BNPR_hybrid, exp_traj)$avg, envelope_new(sim_MCMC, exp_traj)$avg)
mse <- c(mse(BNPR_sim_exp, exp_traj)$avg, mse(BNPR_block_size, exp_traj)$avg, 
         mse(BNPR_hybrid, exp_traj)$avg, mse(sim_MCMC, exp_traj)$avg)

perf_meas <- cbind(coverage, bias, dev, mse)
rownames(perf_meas) <- c('true', 'block_size_mle', 'hybrid_opt', 'MCMC')
perf_meas <- as.data.frame(perf_meas) %>% tibble::rownames_to_column(var = "method") %>%
  pivot_longer(cols=c('coverage', 'bias', 'dev', 'mse'))
print(perf_meas)
```


## Beyond the Beta-coalescent 

We can also reconstruct the effective population size trajectory of a tree under any Lambda-coalescent by specifying the characteristic measure. The following code reconstructs Figure 9 in our paper. 

```{r}
data(NY_flu)
plot(NY_flu, show.tip.label=FALSE)
axisPhylo()

lambda_pmf <- list(values = c(0.2, 0.7, 0.9,1), probs= c(0.25,0.25,0.25,0.25))
lambda_trunc_norm <- function(x) {dtruncnorm(x, a=0, b=1, mean = 0, sd = 1)}

BNPR_Beta_1.5_NY_flu <- BNPR_Lambda(NY_flu, alpha_multif = 1.5) 
BNPR_Lambda_pmf_NY_flu <- BNPR_Lambda(NY_flu, alpha_multif =NULL, lambda_meas = lambda_pmf) 
BNPR_binary_NY_flu <- BNPR(NY_flu)

lambda_trunc_norm <- function(x) {dtruncnorm(x, a=0, b=1, mean = 0, sd = 1)}
BNPR_trunc_norm_NY_flu <- BNPR_Lambda(NY_flu, alpha_multif =NULL, lambda_meas = lambda_trunc_norm) 

par(mfrow=c(2,2))
plot_BNPR(BNPR_binary_NY_flu, main='Kingman Coal')
plot_BNPR(BNPR_Beta_1.5_NY_flu, main=expression(paste('Beta Coal ', alpha, '=1.5')))
plot_BNPR(BNPR_Lambda_pmf_NY_flu, main='Discrete measure')
plot_BNPR(BNPR_trunc_norm_NY_flu, main='Truncated Normal')
```

 
