---
title: "Cas12 daisy data"
author: "Mackenzie Simper, Julia Palacios"
date: "2024-10-10"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cas12 daisy data}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This vignette shows a case study of single lineage tracing after 7 days. The input datafile "/data/d7_rep1". We first only keep the alleles with higher frequency per cell. In this case we have 86 cells.

```{r}
library(phylodyn)
library(tidyr)
library(dplyr)
library(stringr)
library(hash)
library(expm)
library(parallel)
library(phangorn)
data1<-read.csv("https://raw.githubusercontent.com/JuliaPalacios/phylodyn/refs/heads/master/data/d7_rep1",sep=",",header=TRUE)

#For each (cellBC, intBC), multiple alleles can be observed
#We take the most common occurrence (of the entire allele -- all sites together)
data1 <- data1 %>%
  mutate(allele = paste(r1, r2, r3, r4)) %>% #concatenate sites together to get overall allele
  group_by(cellBC, intBC,  lineageGrp) %>%
  summarise(allele_new = names(sort(table(allele),decreasing=TRUE)[1])) %>% #this code finds the most common allele in the group
  mutate(r1 = strsplit(allele_new, split = " ")[[1]][1], r2 = strsplit(allele_new, split = " ")[[1]][2], #split apart into sites again
         r3 = strsplit(allele_new, split = " ")[[1]][3], r4 = strsplit(allele_new, split = " ")[[1]][4]) %>%
  filter(!grepl("NC", allele_new, fixed = TRUE)) #3 entries have NC only in one site, I think it's like NA??
print(nrow(data1))

```

We now transform the data into the format needed for our method as follows

```{r}
S<-4 #number of target sites
data1a<-reformat_slt(data1,S)
```

data1a now consists of 86 cells at 4 target sites with some newly coded "allelles". We will now estimate mutation rates:

```{r}
t<-7 #on the conservative side, this could as well be 6
mu_hat<-estimate_mu(data1a[[1]],t)

print(mu_hat)
print(mu_hat)
           [,1]        [,2]       [,3]       [,4]
[1,] 0.01969114 0.002112431 0.00000000 0.01613936
[2,] 0.00000000 0.049462075 0.00000000 0.00000000
[3,] 0.00000000 0.000000000 0.02871632 0.00000000
[4,] 0.00000000 0.000000000 0.00000000 0.10592879

```

For this dataset, we estimate positive rates for single sites and cross-sites 1-2 and 1-4. All other rates are estimated to be 0.

```{r}
thetas<-list()
thetas[[1]]<-mu_hat

all_estimates<-inference_all(Dlist=data1a, t0=t, thetas=thetas, M = 20)
plot(cumsum(coalescent.intervals(mytree)$interval.length),cumsum(times_estimates$mle_times))

```




Estimation of Ne
==============================================

```{r}
bound<-t
Ne_est<-MLE_BoundCoal(data=all_estimates$tree,lengthout=4,bound=t,eps=.02,eta=.01)
Ne_oracle<-BNPR(mytree,lengthout=4)
plot_BNPR(Ne_oracle)
points(Ne_est$x,Ne_est$effpop,type="S",col="red")
```