---
title: Estimation of an exponential population size trajectory by sampling Tajima’s
  trees (n=10)
author: "Lorenzo Cappello"
date: "30/12/2019"
output: html_document
---

This vignette displays a simulation study designed to approximate an exponential growth population size by sampling Tajima's trees, using `phylodyn`.

We start by loading the required packages.

```{r message=FALSE}
library(phylodyn)
library("rPython")
library("phangorn")
library("mvtnorm")
library("ape")
```

We then set seed and sample size (n), mutation rate (mu), and a scaling factor (fact, set to 1 by default)
```{r}
set.seed(100+8)
n<-10
mu<-75
fact<-1
```


We define an arbitrary population size trajectory (exponential for the purpose of this vignette)
```{r}
bottleneck_traj<-function(t){
  result<-25*exp(-5*t)
  return(result)
}
```

1) SIMULATE DATA

Data simulation proceeds in several steps

We simulate a vector of n-1 coalescent times under the N(t) trajectory defined above
```{r}
simulation<-coalsim(samp_times =0, n_sampled = n, traj = bottleneck_traj)
```

We then simulatue a genealogy with branch lengths defined by the coalescent times simulated above. We also compute the corresponding F matrix representation
```{r}
tree1<-rcoal(n,br=simulation$intercoal_times)
truthF<-create_F(tree1)
```

Last step of the data simulation is to superimpose mutations on the sampled genealogy
```{r}
data1<-simulate_data(mu,tree1)
```

We can check how many mutations we have sampled simply by 
```{r}
nrow(data1)
```

2) INITIALIZATION

Second steps is initializing the MCMC

We initialize the coalescent times and the effective population size

```{r}
initial<-initial_tajima(data1,name="Exp",mu=75,npoints=49,fact=1)
```

Then we initialize the MCMC states
```{r}
setupchain<-initial_MCMC(initial,ngrid=50,mu=75,n=10,alpha=.1,fact=fact)
```

Fix a number of  input variables for MCMC. 

```{r}
acp1<-0;acp2<-0;acp3<-0; acp2_t<-0
currentval<-setupchain$currentval
theta_list<-setupchain$theta_list
prec_list<-setupchain$prec_list
probs_list<-setupchain$probs_list
times_list<-setupchain$times_list
F_list<-setupchain$F_list
```

3) MCMC INFERENCE

Fix the number of MCMC iterations
```{r}
Nsim<-101
```
Start the MCMC iteration

```{r}
for (j in 1:Nsim){
  # Effective population size update
  thetaup<-updateTheta(currentval,theta_list,prec_list,probs_list,const=1,j=1,nsites=mu,eps=.06)
  currentval<-thetaup$currentval; theta_list<-thetaup$theta_list; prec_list<-thetaup$prec_list; probs_list<-thetaup$probs_list
  acp1<-acp1+thetaup$acp

  #Hamiltonian time update
  timesup<-updateTimes(currentval,times_list,theta_list,prec_list,probs_list,const=1,nsites=mu,eps=0.007)
  currentval<-timesup$currentval;theta_list<-timesup$theta_list; prec_list<-timesup$prec_list; probs_list<-timesup$probs_list; times_list<-timesup$times_list
  acp2<-acp2+timesup$acp

  #Local time update
  timesup2<-updateSingleTime(currentval,theta_list,prec_list,probs_list,const=1,nsites=mu)
  currentval<-timesup2$currentval; probs_list<-timesup2$probs_list; times_list<-timesup2$times_list
  acp2_t<-acp2_t+timesup2$acp
  
  #Genealogy update
  result<-currentval$result
  FMatupMarkov_FREE<-updateFmat_Markov_FREE(currentval,F_list,probs_list,const=1,nsites=mu,initial$oldsuff,n)
  currentval<-FMatupMarkov_FREE$currentval;F_list<-FMatupMarkov_FREE$F_list; probs_list<-FMatupMarkov_FREE$probs_list; 
  acp3<-acp3+FMatupMarkov_FREE$acp
  
  if(j %% 100==0) {
      # Print iteration
    print(j)
   }
  

}
```

Save all the output in a list 
```{r}
results<-list(acp1=acp1, acp2=acp2,acp3=acp3,currentval=currentval,times_list=times_list,probs_list=probs_list,theta_list=theta_list,prec_list=prec_list)
```

4) POSTERIOR ANALYSIS

Here we want to plot the posterior of the effective population size. 

We first extract time and log population size MCMC samples

```{r}
times_list=results$times_list
theta_list=results$theta_list
```

And the last value of the chain

```{r}
currentval=results$currentval
```

We then subsample to account for the burnin and subsample frequency chosen 
```{r}
times_list2<-subsample(times_list,burnin=1,subsample=1)
theta_list2<-subsample(theta_list,burnin=1,subsample=1)
```

Prepare the R object to plot
```{r}
res1a<-BNPR(data=tree1,lengthout=100,prec_alpha=.1)
fmed = apply(exp(theta_list2), MARGIN = 1, FUN = stats::median,na.rm=TRUE)
flow = apply(exp(theta_list2), MARGIN = 1, FUN = function(x) stats::quantile(x, .025, na.rm=TRUE) )
fhi  = apply(exp(theta_list2), MARGIN = 1, FUN = function(x) stats::quantile(x, .975, na.rm=TRUE))
intl<-currentval$grid[2]-currentval$grid[1]
grid2<-cumsum(c(0,rep(intl,nrow(theta_list2))))
res_BNPR3<-res1a
fact<-1
midpts2<-grid2[-1]-intl/2
res_BNPR3$x<-midpts2/fact
res_BNPR3$grid<-grid2/fact
res_BNPR3$effpop<-fmed[-length(fmed)]/fact
res_BNPR3$effpop975<-fhi[-length(fmed)]/fact
res_BNPR3$effpop025<-flow[-length(fmed)]/fact
fact<-1
fact2<-1
x<-seq(0,.8,by=.0001)
```

```{r, fig.width=8.5, fig.height=5}
plot_BNPR(res_BNPR3,heatmaps=FALSE,heatmap_labels = FALSE,ylab="Ne(t)",main="(A) Ne(t) BESTT",xlim=c(.77,0),ylim=c(.001,10000))
points(x,bottleneck_traj(x),type="l",lty=2)
```