---
title: "Simulated data - isochronous and heterochronous"
author: "Julia Palacios"
date: "April 13, 2018"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
Data simulation
========================================================

We will first use \texttt{coalsim} to simulate a genealogy from an exponential size trajectory with 10 tips. In all our simulations, time is measured in units of $N_{0}$ generations.

gene = coalsim(samp_times = samp_times, n_sampled = n_sampled, traj = traj, lower_bound = 1/20)

```{r}
library("phylodyn")
library("ape")
library("reticulate")
set.seed(100+8)
n<-10

exp_traj<-function(t){
  result=rep(0,length(t))
  result[t<=0.1]<-10
  result[t>0.1]<-10*exp(2-20*t[t>0.1]) #50 times smaller
  return(result)
}


simulation<-coalsim(samp_times =0, n_sampled = n, traj = exp_traj)
tree1<-rcoal(n,tip.label=NULL,br=simulation$intercoal_times)
```
If we had the genealogy alone, we could use \texttt{phylodyn} to infer $N(t)$ as follows:

```{r}
res1a<-BNPR(data=tree1,lengthout=50,prec_alpha=.1)
#par(mfrow=c(1,2))
plot_BNPR(res1a,heatmaps=FALSE,heatmap_labels = FALSE,ylab="Ne(t)",main="Effective Pop Size",ylim=c(.001,1000))
points(seq(0,sum(simulation$intercoal_times)+.1,by=.01),exp_traj(seq(0,sum(simulation$intercoal_times)+.1,by=.01)),type="l",lty=2)

```

We now simulate mutations along the tree to generate a dataset according to the infinite sites mutation model. 
```{r}
data1<-simulate_data(40,tree1)
nrow(data1)
```
```{r}

constant<-function(x){
  return (rep(1,length(x)))
}

simulation1<-coalsim(0,10,constant)
tree1<-generate_newick(simulation1)
plot(tree1$newick)
tree1$newick
```

In the following example we simulate a genealogy with constant effective population size with 10 tips at time 0 and 40 other tips with uniform sampling times 
```{r}

samp_times = c(0, sort(runif(40, 0, 8)))
n_sampled = c(10, rep(1, 40))

simulation2<-coalsim(samp_times = samp_times, n_sampled = n_sampled, traj = constant)
tree2<-generate_newick(simulation2)
plot(tree2$newick,show.tip.label = F)
tree2$newick

```

Here, we consider a more general demographic model such as bottleneck. We specify our trajectory through the function \texttt{bottleneck_traj} and simulate via the thinning method (see Palacios and Minin, 2013)
```{r}

bottleneck_traj<-function(t){
  result=rep(0,length(t))
  result[t<=0.5]<-1
  result[t>0.5 & t<1]<-.1
  result[t>=1]<-1
  return(result)
}

simulation3<-coalsim(samp_times = samp_times, n_sampled = n_sampled, traj = bottleneck_traj,method="thin",val_upper=11)
tree3<-generate_newick((simulation3))
plot(tree3$newick,show.tip.label = F)
```

Note: Other R packages such as Ape and phyclust (ms) have functions that simulate isochronous genealogies under the coalescent model for specific demographic scenarios such as exponential growth and piece-wise constant functions. 





