---
title: "Plotting perfect phylogeny"
author: "Lorenzo Cappello"
date: "16/4/2023"
output: html_document
---



```{r}
library("phylodyn")
library("seqinr")
library("igraph")


path <- paste(system.file(package="phylodyn"), "data/synth_hetdata.fasta", sep="/")
data<-read.fasta(path)
names<-names(data)
seq_mat<-matrix(unlist(data),nrow=length(data),byrow=TRUE)
```

Remove missing entries (it is enough to check which entries are missing in the first sequence)
```{r,}
id_0<-which(seq_mat[1,]==" ")
seq_mat<-seq_mat[,-id_0]
```


Extract sequence names and sampling times from the fasta label. Sequences 
In this example we have <sequence name>_<sampling time>
```{r,}
seq_names<-c()
samp_times<-c()
sequences<-list()
for (r in names){
    seq_names<-c(seq_names,strsplit(r,"_")[[1]][1])
    samp_times<-c(samp_times,as.numeric(strsplit(r,"_")[[1]][2]))
}
name_samp<-cbind(samp_times,seq_names)
```



Convert data into a 0-1 incidence matrix. In this case, we know that the last sequence is the reference sequence (check the seq_names vector). Otherwise one can upload it from a different source.
```{r,}
inc_mat<-matrix(0,dim(seq_mat)[1],dim(seq_mat)[2])


for (i in 1:dim(inc_mat)[2]){
  inc_mat[1:(dim(inc_mat)[1]-1),i]<-(seq_mat[1:(dim(inc_mat)[1]-1),i]!=seq_mat[dim(inc_mat)[1],i])*1
}

rownames(inc_mat)<-seq_names
```


Now, we put the data in the format that are read by the function in the package:each row represents a segregating site, each column is a sample. 

```{r,}
inc_mat1<-inc_mat[sample(33),]
data_taj<-"0"
for (i in 1:dim(inc_mat)[2]){
  data_taj<-rbind(data_taj,paste(inc_mat1[,i],collapse = ""))
}
data_taj<-as.matrix(data_taj[-1,])
```

Compute the perfect phylogeny and plot it.

```{r,}
n <- as.vector(table(samp_times))
samp_times <- unique(samp_times)
input_tree <- c()
initial<-initial_tajima_het(data_taj,name="trial",mu=1,npoints=49,fact=1,alpha=0.2,samp_times,n,input_tree,name_samp)
oldsuff<-phylodyn:::pp_heterochronous(n_sampled=n,initial$oldsuff,initial$times$addtimes)

#oldsuff<-tajima_perphylo(data_taj)
#Plot perfect phylogeny
mutations<-rep("AAAA",length(oldsuff$mylist))
for (j in 1:length(oldsuff$mylist)){
  mutations[j]<-paste0(oldsuff$mylist[[j]]$y,collapse="-")
}
graph_gene_tree2(oldsuff$nodes,mutations)
```

