---
title: Estimation of Covid population size trajectory by sampling Tajima’s
  trees (Germany)
author: "Lorenzo Cappello"
date: "03/26/2020"
output: html_document
---

This vignette displays a simulation study designed to approximate an exponential growth population size by sampling Tajima's trees, using `phylodyn`.

We start by loading the required packages.

```{r message=FALSE}
library(phylodyn)
library("rPython")
library("phangorn")
library("mvtnorm")
library("ape")
library("truncnorm")
```

Load python function file
```{r}
path <- paste(system.file(package="phylodyn"), "python/Fsamp_hTDv2.py", sep="/")
python.load(path)
```

We first read the observed data in fasta format.

```{r}
library(seqinr)
alig = read.FASTA("~/Documents/Covid_Analysis/alignment/BEAST/fastaGermany.fasta")
n = length(alig)
ids = names(alig)
dates_samp<-ymd(substr(ids, start = nchar(ids) - 9, stop = nchar(ids)))

##A rough estimate of mutation rate
x<-dates_samp-dates_samp[length(dates_samp)]
hamming<-as.matrix(dist.hamming(as.phyDat(alig),ratio=FALSE))
reg<-lm(hamming[nrow(hamming),-nrow(hamming)]~-1+x[-nrow(hamming)])
summary(reg)
plot(x[-nrow(hamming)],hamming[nrow(hamming),-nrow(hamming)],ylab="Distance to reference",xlab="Time difference (Days)",main=paste("A total of ",round(reg$coefficients[[1]]*28,2)," of mutations per month",sep=""))
abline(reg)



#Convert data to 0s and 1s - Assumes the reference sequence is the last row
data.matrix<-as.character(as.matrix(alig))
mu<-reg$coefficients[[1]]*365
tot.row<-nrow(data.matrix)
binary.matrix<-matrix(0,nrow=nrow(data.matrix),ncol=ncol(data.matrix))
for (j in 1:ncol(binary.matrix)){
  tmp<-data.matrix[data.matrix[,j]!=data.matrix[tot.row,j],j]
  tmp<-tmp[tmp!="n"]
  tmp<-tmp[tmp!="-"]
  tmp2<-unique(c(tmp[tmp=="a"],tmp[tmp=="c"],tmp[tmp=="t"],tmp[tmp=="g"]))
  if (length(tmp2)==1){binary.matrix[data.matrix[,j]==tmp2[1],j]<-1}
  if (length(tmp2)==2){
    if (min(dates.tot[data.matrix[,j]==tmp2[1]])<min(dates.tot[data.matrix[,j]==tmp2[2]])){
      binary.matrix[data.matrix[,j]==tmp2[1],j]<-1
      binary.matrix[data.matrix[,j]==tmp2[2],j]<-2
    }else{
      binary.matrix[data.matrix[,j]==tmp2[1],j]<-2
      binary.matrix[data.matrix[,j]==tmp2[2],j]<-1
    }
    
  }
  if (length(tmp2)==3){
    print(j)
    break
  }
  
}

tmp<-apply(binary.matrix,2,sum)
#Only polymorphic sites
binary.matrix<-binary.matrix[,tmp>0]
##make it in Tajima Format
samp.times<-decimal_date(dates_samp)
order.date<-sort(samp.times,index.return=TRUE,decreasing = TRUE)
order.date$x
binary.matrix<-binary.matrix[order.date$ix,]

datatajima<-paste0(binary.matrix[,1],collapse="",sep="")
for (j in 2:ncol(binary.matrix)){
  datatajima<-rbind(datatajima,paste0(binary.matrix[,j],collapse="",sep=""))
}

labels<-ids[order.date$ix]
ntab<-rev(table(samp.times))
samp_times<-max(samp.times)-sort(unique(samp.times),decreasing=TRUE)
```

We can check how many mutations we have sampled simply by 
```{r}
length(datatajima)
```

2) INITIALIZATION

Second steps is initializing the MCMC

We initialize the genealogy(ranked tree shape + coalescent times) and the effective population size.
We also create a file converting the data into nucleotide basis. 
It is possible either to rely on the serial UPGMA initial genealogy (option 1) or include and \texttt{input_tree} in Newick format (option 2). \texttt{name_samp} is a an array of two columns and number of rows equal to the number of samples, in which the first column is the sampling time, the second the label in the input genealogy. In option 1, the \texttt{input_tree} and \texttt{name_samp} are left blank
```{r}
set.seed(123)
input_tree<-c()
name_samp<-c()
fact<-1
initial<-initial_tajima_het(datatajima,name="exp70",mu,npoints=49,fact=1,alpha=0.2,samp_times,ntab,input_tree,name_samp)
```

Then we initialize the MCMC states
```{r}
setupchain<-initial_MCMC_het(initial,ngrid=50,mu,ntab,alpha=.1,fact,Nate=100,initial$times,samp_times)
```

Set inputs variables for the MCMC (eps:step size HMC, Z1 and sigma see Cappello et al. 2020 Section 4,NSim: lenght of the MC chain)
```{r}
eps<-0.07
Z1<-2
sigma<-0.02
Nsim<-300
```

Set a number of arrays to store ouput of the MCMC
```{r}
acp1<-0;acp2<-0;acp3<-0; acp2_t<-0
currentval<-setupchain$currentval
theta_list<-setupchain$theta_list
prec_list<-setupchain$prec_list
probs_list<-setupchain$probs_list
times_list<-setupchain$times_list
oldsuff<-initial$oldsuff
```

3) MCMC INFERENCE


Start the MCMC iteration



```{r}
n<-ntab
for (j in 1:Nsim){
  #1)Hamiltionian MC update for the effective population size (Lan et al. 2015)
  thetaup<-updateTheta_het(currentval,theta_list,prec_list,probs_list,const=1,1,nsites=mu,eps)
  thetaup$acp
  currentval<-thetaup$currentval; theta_list<-thetaup$theta_list; prec_list<-thetaup$prec_list; probs_list<-thetaup$probs_list
  acp1<-acp1+thetaup$acp

  #2) Local-Global update for the coalescent times via a truncated normal (Cappello et al. 2020)
  timesup<-updateTimes_hetLocal(n,oldsuff,currentval,times_list,theta_list,prec_list,probs_list,const=1,mu,samp_times,Z1,sigma)
  currentval<-timesup$currentval;theta_list<-timesup$theta_list; prec_list<-timesup$prec_list; probs_list<-timesup$probs_list; times_list<-timesup$times_list
  timesup$acp
  acp2<-acp2+timesup$acp

  #3) Local update for the ranked tree shapes (Palacios et al. 2019, Cappello et al. 2020)
  FMatupMarkov_FREE<-updateFmat_Markov_FREE_het(currentval,probs_list,const=1,mu,oldsuff,n)
  currentval<-FMatupMarkov_FREE$currentval; probs_list<-FMatupMarkov_FREE$probs_list;
  acp3<-acp3+FMatupMarkov_FREE$acp



}
```

Save all the output in a list 
```{r}
results<-list(acp1=acp1, acp2=acp2,acp3=acp3,currentval=currentval,times_list=times_list,probs_list=probs_list,theta_list=theta_list,prec_list=prec_list)
```

4) POSTERIOR ANALYSIS

Here we want to plot the posterior of the effective population size. 

We first extract time and log population size MCMC samples

```{r}
times_list=results$times_list
theta_list=results$theta_list
```

And the last value of the chain

```{r}
currentval=results$currentval
```

We then subsample to account for the burnin and subsample frequency chosen 
```{r}
times_list2<-subsample(times_list,burnin=1,subsample=1)
theta_list2<-subsample(theta_list,burnin=1,subsample=1)
```

Prepare the R list to plot the effective population size
```{r}
res_BNPR3<-initial$res2b
fmed = apply(exp(theta_list2), MARGIN = 1, FUN = stats::median,na.rm=TRUE)
flow = apply(exp(theta_list2), MARGIN = 1, FUN = function(x) stats::quantile(x, .025, na.rm=TRUE) )
fhi  = apply(exp(theta_list2), MARGIN = 1, FUN = function(x) stats::quantile(x, .975, na.rm=TRUE))
intl<-currentval$grid[2]-currentval$grid[1]
grid2<-cumsum(c(0,rep(intl,nrow(theta_list2))))
fact<-1
midpts2<-grid2[-1]-intl/2
res_BNPR3$x<-midpts2/fact
res_BNPR3$grid<-grid2/fact
res_BNPR3$effpop<-fmed[-length(fmed)]/fact
res_BNPR3$effpop975<-fhi[-length(fmed)]/fact
res_BNPR3$effpop025<-flow[-length(fmed)]/fact
```

```{r, fig.width=8.5, fig.height=5}
plot_BNPR(res_BNPR3,heatmaps=FALSE,heatmap_labels = FALSE,ylab="Ne(t)",main="(A) Ne(t) BESTT",ylim=c(.00001,4),xlim=c(.25,0))
```
